---
title: "Computer Work"
subtitle: null
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    highlight: tango
    theme: cosmo
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# Okun's law

Professor House mentioned the "statistical pattern" known as Okun's law in class.
From the "Lecture_2_handouts":
$$
\% \Delta \text{GDP} \approx 3 \% - 2 \% \times \Delta \text{unemployment rate}.
$$
Plotting this relationship involves two steps.
We first need to grab the data from FRED.
Second, we need to regress the percent change in GDP against the change in the unemployment rate to uncover estimates of the two above parameters that define the linear relationship.

## Getting the data from FRED

```{r}
library(tidyverse)
library(lubridate)
library(tidyquant)
dat <- tq_get(c("UNRATE", "GDPC1"), get = "economic.data")
# "Spread" the data
dat <- dat %>% 
  spread(key = symbol, value = price) %>% 
  mutate(year = year(date)) %>% 
  select(-date)
# Get annual averages
by_year <- group_by(dat, year)
dat <- summarise_all(by_year, mean, na.rm = TRUE)
head(dat)
tail(dat)
```


## Estimating the pattern

# Real interest rates from homework 1

Read in the data that is provided on Canvas:
```{r}
library(tidyverse)
library(lubridate)
dat <- readxl::read_excel("../PS1-data_2018.xlsx")
head(dat)
tail(dat)
```
Next we want to compute the inflation rate as a percent change at an annual rate.
The formula for computing a compounded annual rate of change is 
$$
\left[ \left( \frac{x_t}{x_{t-1}} \right)^{\text{number obs per year}} - 1 \right] \times 100,
$$
where $x_t$ is the value of the series at time $t$.
For monthly data the number of observations per year is $12$.
A useful overview of similar transformations is available at the FRED [website](https://fredhelp.stlouisfed.org/fred/data/understanding-the-data/formulas-calculate-growth-rates/).
To accomplish this in R we can use the `lag` command:

```{r}
dat <- dat %>% 
  mutate(LCPIU = lag(CPIU))
head(dat)
tail(dat)
dat <- dat %>% 
  mutate(infl = 100*((CPIU / LCPIU)^12 - 1),
         datem = ymd(paste0(Year, "-", Month, "-01")))

dat <- dat %>%
  filter(datem <= ymd("2015-11-01"),
         datem >= ymd("1960-01-01"))
ggplot(data = dat) +
  geom_line(mapping = aes(x = datem, y = infl)) +
  xlab("") + ylab("Percent per year") + ggtitle("Inflation")
```

Now we can compute the _ex post_ real interest rate:
$$
r = FEDFUNDS - infl.
$$
```{r}
dat <- dat %>% 
  mutate(r = FEDFUNDS - infl)
```
And plot the resulting series:
```{r}
ggplot(data = dat) +
  geom_line(mapping = aes(x = datem, y = r)) +
  xlab("") + ylab("Percent per year") + ggtitle("Real interest rates")
```

```{r}
avg <- mean(dat$r, na.rm = TRUE)
```

The real interest rate in `r max(dat$datem)` is `r dat$r[[which(dat$datem == max(dat$datem))]]`.
The average real interest rate over the sample is `r avg`.

Ideally you'd like to list those numbers using R code.
Here's the code I used: `max(dat$datem)` and `dat$r[[which(dat$datem == max(dat$datem))]]`.
Putting an `r` in from of those commands lists the output within the R markdown,
resulting in the output you see above.


We can add recession shading and the sample avarage:
```{r}
recessions_df <- read.table(textConnection(
"Peak, Trough
1857-06-01, 1858-12-01
1860-10-01, 1861-06-01
1865-04-01, 1867-12-01
1869-06-01, 1870-12-01
1873-10-01, 1879-03-01
1882-03-01, 1885-05-01
1887-03-01, 1888-04-01
1890-07-01, 1891-05-01
1893-01-01, 1894-06-01
1895-12-01, 1897-06-01
1899-06-01, 1900-12-01
1902-09-01, 1904-08-01
1907-05-01, 1908-06-01
1910-01-01, 1912-01-01
1913-01-01, 1914-12-01
1918-08-01, 1919-03-01
1920-01-01, 1921-07-01
1923-05-01, 1924-07-01
1926-10-01, 1927-11-01
1929-08-01, 1933-03-01
1937-05-01, 1938-06-01
1945-02-01, 1945-10-01
1948-11-01, 1949-10-01
1953-07-01, 1954-05-01
1957-08-01, 1958-04-01
1960-04-01, 1961-02-01
1969-12-01, 1970-11-01
1973-11-01, 1975-03-01
1980-01-01, 1980-07-01
1981-07-01, 1982-11-01
1990-07-01, 1991-03-01
2001-03-01, 2001-11-01
2007-12-01, 2009-06-01"), sep=',', 
colClasses=c('Date', 'Date'), 
header=TRUE)

recessions_trim <- subset(recessions_df, Peak >= min(dat$datem))
ggplot(data = dat) +
  geom_line(mapping = aes(x = datem, y = r)) +
  geom_hline(yintercept = avg, color = "red") +
  geom_rect(data = recessions_trim, 
            aes(xmin = Peak, xmax = Trough,
                ymin = -Inf, ymax = +Inf),
            fill = "pink", alpha = 0.3) +
  xlab("") + ylab("Percent per year") + 
  ggtitle("Real interest rates", subtitle = "with recession shading")
```
 
 