---
title: "Computer Work"
subtitle: null
output:
  html_document:
    highlight: tango
    theme: cosmo
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# Okun's law

Professor House mentioned the "statistical pattern" known as Okun's law in class.
From the "Lecture_2_handouts":
$$
\% \Delta \text{GDP} \approx 3 \% - 2 \% \times \Delta \text{unemployment rate}.
$$
Plotting this relationship involves two steps.
We first need to grab the data from FRED.
Second, we need to regress the percent change in GDP against the change in the unemployment rate to uncover estimates of the two above parameters that define the linear relationship.
The R code replicates figure 10-4 in Mankiw, 9e, p 286.

## Getting the data from FRED

The first step is getting the data from FRED and
creating annual averages of the data.
```{r}
library(tidyverse)
library(lubridate)
library(tidyquant)
dat <- tq_get(c("UNRATE", "GDPC1"), 
              get = "economic.data",
              from = "1940-01-01")
# "Spread" the data
dat <- dat %>% 
  spread(key = symbol, value = price) %>% 
  mutate(year = year(date)) %>% 
  select(-date)
# Get annual averages
by_year <- group_by(dat, year)
dat <- summarise_all(by_year, mean, na.rm = TRUE)
head(dat)
tail(dat)
```
We're interested in the percent change in real GDP and the difference in the unemployment rates:
```{r}
# Calculate percent change in rGDP and the difference in the unemployment rate
dat <- dat %>% 
  mutate(gdp = 100*(GDPC1 / lag(GDPC1) - 1),
         ur = UNRATE - lag(UNRATE))
```


## Estimating the pattern

At this point we're ready to estimate the relationship between the percentage change in real GDP and the change in the unemployment rate posited by Okun:
```{r}
# Run the regression
okun <- lm(gdp ~ ur, data = dat)
print(okun)
```
With rounding, we're at the coefficients mentioned in class,
which is a minor victory.

The last step is plotting the data.
To do so, we need to add the regression line:
```{r}
xur <- seq(-2, 3)
reg_line <- okun$coefficients[[1]] + okun$coefficients[[2]]*xur
fit <- tibble(xur, reg_line)

ggplot(data = dat) +
  geom_point(mapping = aes(x = ur, y = gdp)) +
  geom_line(mapping = aes(x = xur, y = reg_line), 
            data = fit,
            color = "blue", 
            size = 1.2) +
  xlab("Change in the unemployment rate") + 
  ylab("Percentage change in real GDP") +
  ggtitle("Okun's law")
```

Just fyi:
you can add a regression line in R without having to run the regression using R's built-in commands.
I did these calculations by hand just for practice.

# Real interest rates from homework 1

Read in the data that is provided on Canvas:
```{r}
library(tidyverse)
library(lubridate)
dat <- readxl::read_excel("../PS1-data_2018.xlsx")
head(dat)
tail(dat)
```
Next we want to compute the inflation rate as a percent change at an annual rate.
The formula for computing a compounded annual rate of change is 
$$
\left[ \left( \frac{x_t}{x_{t-1}} \right)^{\text{number obs per year}} - 1 \right] \times 100,
$$
where $x_t$ is the value of the series at time $t$.
For monthly data the number of observations per year is $12$.
A useful overview of similar transformations is available at the FRED [website](https://fredhelp.stlouisfed.org/fred/data/understanding-the-data/formulas-calculate-growth-rates/).
To accomplish this in R we can use the `lag` command:

```{r}
dat <- dat %>% 
  mutate(LCPIU = lag(CPIU))
head(dat)
tail(dat)
dat <- dat %>% 
  mutate(infl = 100*((CPIU / LCPIU)^12 - 1),
         datem = ymd(paste0(Year, "-", Month, "-01")))

dat <- dat %>%
  filter(datem <= ymd("2015-11-01"),
         datem >= ymd("1960-01-01"))
ggplot(data = dat) +
  geom_line(mapping = aes(x = datem, y = infl)) +
  xlab("") + ylab("Percent per year") + ggtitle("Inflation")
```

Now we can compute the _ex post_ real interest rate:
$$
r = FEDFUNDS - infl.
$$
```{r}
dat <- dat %>% 
  mutate(r = FEDFUNDS - infl)
```
And plot the resulting series:
```{r}
ggplot(data = dat) +
  geom_line(mapping = aes(x = datem, y = r)) +
  xlab("") + ylab("Percent per year") + ggtitle("Real interest rates")
```

```{r}
avg <- mean(dat$r, na.rm = TRUE)
```

The real interest rate in `r max(dat$datem)` is `r dat$r[[which(dat$datem == max(dat$datem))]]`.
The average real interest rate over the sample is `r avg`.

Ideally you'd like to list those numbers using R code.
Here's the code I used: `max(dat$datem)` and `dat$r[[which(dat$datem == max(dat$datem))]]`.
Putting an `r` in from of those commands lists the output within the R markdown,
resulting in the output you see above.


We can add recession shading and the sample avarage:
```{r}
recessions_df <- read.table(textConnection(
"Peak, Trough
1857-06-01, 1858-12-01
1860-10-01, 1861-06-01
1865-04-01, 1867-12-01
1869-06-01, 1870-12-01
1873-10-01, 1879-03-01
1882-03-01, 1885-05-01
1887-03-01, 1888-04-01
1890-07-01, 1891-05-01
1893-01-01, 1894-06-01
1895-12-01, 1897-06-01
1899-06-01, 1900-12-01
1902-09-01, 1904-08-01
1907-05-01, 1908-06-01
1910-01-01, 1912-01-01
1913-01-01, 1914-12-01
1918-08-01, 1919-03-01
1920-01-01, 1921-07-01
1923-05-01, 1924-07-01
1926-10-01, 1927-11-01
1929-08-01, 1933-03-01
1937-05-01, 1938-06-01
1945-02-01, 1945-10-01
1948-11-01, 1949-10-01
1953-07-01, 1954-05-01
1957-08-01, 1958-04-01
1960-04-01, 1961-02-01
1969-12-01, 1970-11-01
1973-11-01, 1975-03-01
1980-01-01, 1980-07-01
1981-07-01, 1982-11-01
1990-07-01, 1991-03-01
2001-03-01, 2001-11-01
2007-12-01, 2009-06-01"), sep=',', 
colClasses=c('Date', 'Date'), 
header=TRUE)

recessions_trim <- subset(recessions_df, Peak >= min(dat$datem))
ggplot(data = dat) +
  geom_line(mapping = aes(x = datem, y = r)) +
  geom_hline(yintercept = avg, color = "red") +
  geom_rect(data = recessions_trim, 
            aes(xmin = Peak, xmax = Trough,
                ymin = -Inf, ymax = +Inf),
            fill = "pink", alpha = 0.3) +
  xlab("") + ylab("Percent per year") + 
  ggtitle("Real interest rates", subtitle = "with recession shading")
```
 
# Consumption around the Great Recession

You explored the motive for smoothing consumption in class.
So you can imagine the shock economists experienced when they saw this picture as the Great Recession unfolded:

```{r}
# Real Personal Consumption Expenditures (PCECC96)	DOWNLOAD 
# Observation: Q3 2017: 11,916.576 (+ more)  
# Updated: Dec 21, 2017
# Units: Billions of Chained 2009 Dollars, Seasonally Adjusted Annual Rate
# Frequency: Quarterly

dat <- tq_get("PCECC96", get = "economic.data", 
              from = "2005-01-01",
              to = "2012-01-01")

recessions_trim <- subset(recessions_df, Peak >= min(dat$date))
ggplot(data = dat) +
  geom_line(mapping = aes(x = date, y = price)) +
  geom_rect(data = recessions_trim, 
            aes(xmin = Peak, xmax = Trough,
                ymin = -Inf, ymax = +Inf),
            fill = "pink", alpha = 0.4) +
  xlab("") + ylab("Billions of real 2009 dollars") + 
  ggtitle("Real personal consumption expenditures")
```

 